---
title: "Machine Learning - Binary Logistic Regression"
author: "Richard Adejumola"
date: "2026-04-10"
output: 
  html_document:
    toc: true
    number_sections: true
    code_folding: show
---



#### Load relevant libraries
```{r, include=FALSE}
library(janitor)
library(skimr)
library(tidyverse)
library(tidymodels)
library(tidytext)
library(tidyclust)
library(ClusterR)
library(klaR)
library(GGally)
library(knitr)
library(gtsummary)
library(devtools)
library(ggmosaic)

library(reportRmd)
library(sjPlot)
library(plotly)
library(psych)
library(parallel)
library(finalfit)
library(gtsummary)
library(mlbench)
library(vip)
library(rsample)
library(tune)
library(recipes)
library(yardstick)
library(parsnip)
library(glmnet)
library(themis)
library(corrr)
library(performance)
library(utils)
library(see)

library(themis)
library(kknn)
library(microbenchmark)
pacman::p_load(
  rio,            # import/export
  here,           # file pathways
  flextable,      # make HTML tables 
  officer,        # helper functions for tables
  tidyverse)      # data management, summary, and visualization
```

# Data Wrangling
## Import survey data

The data comes from the 2024 Behavioral Risk Factor Surveillance System. There were 457,670 respondents and 301 variables. 
```{r}
brfss_data24 <- read_csv("../brfssdata/brfss24.csv")
```


## Filter survey data for relevant variables

We are only interested in a subset of variables that we suspect predict the outcome of lifetime diagnosis of depression among persons with diabetes. Here, we filter and rename variables. 
```{r class.source = 'fold-hide'}
fp_data_vars <- brfss_data24 %>% 
      select(
        id = SEQNO,                         metropolis = '_METSTAT',            marital = MARITAL, 
        state = '_STATE',                   sex = SEXVAR,                       
        bmi = "_BMI5",                      bmi_cat = "_BMI5CAT",               education = '_EDUCAG',
        employment = EMPLOY1,               sight = BLIND,                      age =  '_AGE80',
        age_grp = '_AGE_G',                 ethnicity = "_IMPRACE",             income2 = '_INCOMG1',
        income = INCOME3,                   weight_lb = WEIGHT2,                height_in = HEIGHT3,
        health_rating = GENHLTH,            physical_health = '_PHYS14D',       mental_health = '_MENT14D',         
        memory = DECIDE,                    walk = DIFFWALK,                    deaf = DEAF,                                       health_insurance = PRIMINS2,        exercise = EXERANY2,
        is_insured = '_HLTHPL2',            last_checkup = CHECKUP1,            heart_attack = CVDINFR4,                           coronary_hd = CVDCRHD4,             stroke = CVDSTRK3,                 asthma = ASTHMA3,                                   ever_cancer = CHCOCNC1,             ever_copd = CHCCOPD3, 
        ever_depression = ADDEPEV3,         ever_ckd = CHCKDNY2,                age_had_dm = DIABAGE4,
        ever_arthritis = HAVARTH4,          ever_diabetes = DIABETE4,           dm_type = DIABTYPE,   
        checked_hba1c = CHKHEMO3,           dm_education = DIABEDU1,            dm_feet = FEETSORE,        
        )
```

## Filter survey respondents for the study sample

This data step filters for observations who have diabetes and know the type of diabetes they have. Persons who were not diagnosed with diabetes at the time of survey are excluded from downstream analysis. 

There are 65,809 individuals with diabetes. However, diabetes type was only reported for 13,780. 

```{r}
fp_data <- fp_data_vars %>% 
    mutate(
      ever_diabetes = case_when(
        ever_diabetes > 1 ~ NA_real_, #code as missing
        .default = ever_diabetes
      ))
      
fp_data_sample <- fp_data %>% drop_na('ever_diabetes' ) 
knows_dm_type <- fp_data_sample %>% drop_na('dm_type' ) 

fp_data_sample %>% tabyl(ever_diabetes)
knows_dm_type %>% tabyl(dm_type)

```


#### Randomly assign diabetes type for missing response

If we perform the analysis on 13,780 only, we exclude a large proportion of the data. Hence, we will assign a diabetes type by random assignment to the individuals who reported having diabetes but whose diabetes type was not collected. The assignment is as such: 90% type 1 diabetes and 10% type 2 diabetes. The rationale is T2DM accounts for approximately 90% of all cases of diabetes diagnosed in the United States (and globally). 

```{r}
fp_data_sample %>% tabyl(dm_type)

set.seed(1123581321)

fp_data_sample2 <- fp_data_sample %>%
  mutate(row_id = row_number()) %>%
  
  left_join(
    fp_data_sample %>%
      mutate(row_id = row_number()) %>%
      filter(is.na(dm_type) | dm_type==7) %>% 
      slice_sample(prop = 0.9) %>%
      mutate(to_recode = TRUE) %>%
      select(row_id, to_recode),
    by = "row_id"
  ) %>%
  
  mutate(
    dm_type = case_when(
      is.na(dm_type) & to_recode == TRUE  ~ 2,
      is.na(dm_type) & is.na(to_recode)   ~ 1,
        dm_type == 7 & to_recode == TRUE  ~ 2,
        dm_type == 7 & is.na(to_recode)   ~ 1,
      dm_type == 9 ~ NA_real_, #code refused as missing
      .default = dm_type
    )
  ) %>%
  
  select(-row_id, -to_recode)   

fp_data_sample2 %>% tabyl(dm_type) 
```

#### View summary of imported data
```{r results='hide'}
summary(fp_data_sample2)
```

## Additional Data preprocessing

At this stage, we are considering a new, smaller subset of variables from the ones filtered from the original survey data. 

#### Select candidate variables for the final models
```{r results='hide'}
data_myvars <- fp_data_sample2 %>% 
      select(
        id, sex, metropolis, bmi_cat, marital, education, employment, ethnicity, age_grp, physical_health, mental_health, last_checkup, exercise, heart_attack, stroke, sight, memory, walk, deaf, ever_arthritis, asthma, ever_depression, age_had_dm, dm_type) 

summary(data_myvars)
```

#### Clean variables 

Some levels of variables include invalid levels that should be considered missing. For instance, income, education, employment, and marital status, which all have levels "7", 9", and "11", which are missing response categories in the 2024 BRFSS survey codebook. 

```{r results='hide'}

data_myvars_clean <- data_myvars %>% 
    mutate(
      education = case_when(
        education > 4 ~ NA_real_,
        .default = education
      ),
      employment = case_when(
        employment > 8 ~ NA_real_,
          .default = employment
      ),
      marital = case_when(
        marital > 6 ~ NA_real_,
          .default = marital
      ),
      physical_health = case_when(
        physical_health > 3 ~ NA_real_,
          .default = physical_health
      ),
      mental_health = case_when(
        mental_health > 3 ~ NA_real_,
          .default = mental_health
      ),
      memory = case_when(
        memory > 2 ~ NA_real_,
          .default = memory
      ),
      walk = case_when(
        walk > 2 ~ NA_real_,
          .default = walk
      ),
      deaf = case_when(
        deaf > 2 ~ NA_real_,
          .default = deaf
      ),
      sight = case_when(
        sight > 2 ~ NA_real_,
        .default = sight
      ),
      last_checkup = case_when(
        last_checkup > 4 ~ NA_real_,
          .default = last_checkup
      ),
      exercise = case_when(
        exercise > 2 ~ NA_real_,
          .default = exercise
      ),
      heart_attack = case_when(
        heart_attack > 2 ~ NA_real_,
          .default = heart_attack
      ),
      stroke = case_when(
        stroke > 2 ~ NA_real_,
          .default = stroke
      ), 
      asthma = case_when(
        asthma > 2 ~ NA_real_,
          .default = asthma
      ), 
      ever_depression = case_when(
        ever_depression > 2 ~ NA_real_,
          .default = ever_depression
      ),
      ever_arthritis = case_when(
        ever_arthritis > 2 ~ NA_real_,
          .default = ever_arthritis
      ),
      age_had_dm = case_when(
        age_had_dm > 97 ~ NA_real_,
          .default = age_had_dm
      )
    )

summary(data_myvars_clean)
```



#### Review missingness in candidate variables for final model
```{r}
skim(data_myvars_clean)
data_myvars_clean %>% tabyl(sex)
data_myvars_clean %>% tabyl(metropolis)
data_myvars_clean %>% tabyl(bmi_cat)
data_myvars_clean %>% tabyl(marital)
data_myvars_clean %>% tabyl(ethnicity)
data_myvars_clean %>% tabyl(education)
data_myvars_clean %>% tabyl(employment)

data_myvars_clean %>% tabyl(mental_health)
data_myvars_clean %>% tabyl(physical_health)
data_myvars_clean %>% tabyl(age_grp)
data_myvars_clean %>% tabyl(last_checkup)
data_myvars_clean %>% tabyl(exercise)
data_myvars_clean %>% tabyl(heart_attack)
data_myvars_clean %>% tabyl(asthma)

data_myvars_clean %>% tabyl(stroke)

data_myvars_clean %>% tabyl(ever_depression)

data_myvars_clean %>% tabyl(ever_arthritis)
data_myvars_clean %>% tabyl(dm_type)
data_myvars_clean %>% tabyl(memory)
data_myvars_clean %>% tabyl(walk)
data_myvars_clean %>% tabyl(deaf)
data_myvars_clean %>% tabyl(sight)
data_myvars_clean %>% tabyl(dm_type)
```


#### Drop obs with missing values
```{r results='hide'}
final_data <- data_myvars_clean %>% 
      drop_na(
        id, sex, metropolis, bmi_cat, marital, education, employment, ethnicity, age_grp, physical_health, mental_health, last_checkup, exercise, heart_attack, stroke, sight, memory, walk, deaf, ever_arthritis, asthma, ever_depression, age_had_dm, dm_type) 

summary(final_data)
```

#### Recode reference level of variables
```{r results='hide'}

final_data_recoded <- final_data %>% 
    mutate(
      sex = case_when(
        sex == 2 ~ 1,
        sex == 1 ~ 0, #male ref level
        .default = sex
      ),
      metropolis = case_when(
        metropolis == 1 ~ 0, #yes
        metropolis == 2 ~ 1, 
        .default = metropolis
      ),
      sight = case_when(
        sight == 2 ~ 0, #no ref
        .default = sight
      ),
      memory = case_when(
        memory == 2 ~ 0, #no ref
        .default = memory
      ),
      walk = case_when(
        walk == 2 ~ 0, #no ref
        .default = walk
      ),
      deaf = case_when(
        deaf == 2 ~ 0, #no ref
        .default = deaf
      ),
      bmi_cat = case_when(
        bmi_cat == 2 ~ 0, #normal weight ref level
        bmi_cat == 1 ~ 1,
        bmi_cat == 3 ~ 2,
        bmi_cat == 4 ~ 3,
        .default = bmi_cat
      ),
      marital = case_when(
        marital == 1 ~ 0, #married ref
        marital == 2 ~ 1,
        marital == 3 ~ 1,
        marital == 4 ~ 1, 
        marital == 5 ~ 2,
        marital == 6 ~ 2,
          .default = marital
      ),
      education = case_when(
        education == 4 ~ 0, #college complete ref
        .default = education
      ),
      employment = case_when(
        employment == 1 ~ 0, #employed/self-employed ref.
        employment == 2 ~ 0,
        employment == 3 ~ 1,
        employment == 4 ~ 1,
        employment == 5 ~ 1,
        employment == 6 ~ 1,
        employment == 7 ~ 1,
        employment == 8 ~ 1,
        .default = employment
      ),
       
      ethnicity = case_when(
        ethnicity == 1 ~ 0, #white ref.
        ethnicity == 2 ~ 1, #black
        ethnicity == 3 ~ 2,
        ethnicity == 4 ~ 2, 
        ethnicity == 5 ~ 2,
        ethnicity == 6 ~ 2,
          .default = ethnicity
      ),
      age_grp = case_when(
        age_grp == 6 ~ 0, #65 years and older ref
        age_grp == 1 ~ 1,
        age_grp == 2 ~ 1,
        age_grp == 3 ~ 1,
        age_grp == 4 ~ 1,
        age_grp == 5 ~ 2,
          .default = age_grp
      ),
      physical_health = case_when(
        physical_health == 1 ~ 0, #zero days ref
        physical_health == 2 ~ 1,
        physical_health == 3 ~ 2,
          .default = physical_health
      ),
      mental_health = case_when(
        mental_health == 1 ~ 0, #zero days ref
        mental_health == 2 ~ 1,
        mental_health == 3 ~ 2,
          .default = mental_health
      ),
      last_checkup = case_when(
        last_checkup == 1 ~ 0, #within last year ref
        last_checkup == 2 ~ 1, 
        last_checkup == 3 ~ 1, 
        last_checkup == 4 ~ 1,
        last_checkup == 5 ~ 1,
          .default = last_checkup
      ),
      exercise = case_when(
        exercise == 1 ~ 0, # yes ref
        exercise == 2 ~ 1, 
          .default = exercise
      ),
      stroke = case_when(
        stroke == 2 ~ 0,
          .default = stroke
      ), 
      asthma = case_when(
        asthma == 2 ~ 0,
          .default = asthma
      ), 
      
      heart_attack = case_when(
        heart_attack == 2 ~ 0,
          .default = heart_attack
      ),
      
      ever_depression = case_when(
        ever_depression == 2 ~ 0,
          .default = ever_depression
      ),
      
      ever_arthritis = case_when(
        ever_arthritis == 2 ~ 0,
          .default = ever_arthritis
      ),
     
      dm_type = case_when(
        dm_type == 2 ~ 0,
          .default = dm_type
      )
    )

summary(final_data_recoded)
```

#### Check to see that recoding worked
```{r}
skim(final_data_recoded)
final_data_recoded %>% tabyl(sex)
final_data_recoded %>% tabyl(metropolis)
final_data_recoded %>% tabyl(bmi_cat)
final_data_recoded %>% tabyl(marital)
final_data_recoded %>% tabyl(ethnicity)
final_data_recoded %>% tabyl(education)
final_data_recoded %>% tabyl(employment)

final_data_recoded %>% tabyl(mental_health)
final_data_recoded %>% tabyl(physical_health)
final_data_recoded %>% tabyl(age_grp)
final_data_recoded %>% tabyl(last_checkup)
final_data_recoded %>% tabyl(exercise)
final_data_recoded %>% tabyl(heart_attack)
final_data_recoded %>% tabyl(asthma)

final_data_recoded %>% tabyl(stroke)
final_data_recoded %>% tabyl(ever_depression)
final_data_recoded %>% tabyl(ever_arthritis)

final_data_recoded %>% tabyl(dm_type)
final_data_recoded %>% tabyl(memory)
final_data_recoded %>% tabyl(walk)
final_data_recoded %>% tabyl(deaf)
final_data_recoded %>% tabyl(sight)
```

#### Convert variables to factor
```{r}
diabetes_data_factor <- final_data_recoded %>%
  mutate(across(c(sex, bmi_cat, sight, memory, walk, deaf, metropolis, marital, education, employment, ethnicity, age_grp, physical_health, mental_health, last_checkup, exercise, heart_attack, stroke, asthma, ever_depression, ever_arthritis, dm_type), as.factor))
```

## Data Split 
Split the dataset into training and test sets
```{r}
set.seed(1123581321)

data_split <- initial_validation_split(
  diabetes_data_factor,
  strata = dm_type, 
  prop = c(0.6, 0.05)) 

train_data <- training(data_split)
test_data     <- testing(data_split)

```

# Descriptive Statistics
```{r}
descriptives <- diabetes_data_factor

descriptives <- descriptives %>%
	mutate(depression = case_when(
		ever_depression == 0 ~ "No",
    ever_depression == 1 ~ "Yes"
	),
	dmtype = case_when(
		dm_type == 0 ~ "Type 2 DM",
    dm_type == 1 ~ "Type 1 DM"
	),
	sex = case_when(
		sex == 0 ~ "Male",
    sex == 1 ~ "Female"
	)
	)

#Chart of Depression
bar_depression <- ggplot(descriptives, aes(depression)) + 
                geom_bar() + 
                labs(x = "Proportion of adults living with diabetes with depression") + 
                theme(axis.text.x = element_text(angle = 360))       
plot(bar_depression)

#Chart of DM Type
bar_dmtype <- ggplot(descriptives, aes(dmtype)) + 
                geom_bar() + 
                labs(x = "Type of diabetes") + 
                theme(axis.text.x = element_text(angle = 360))       
plot(bar_dmtype)

# Chart of age of diabetes diagnosis
agedm_histo <- ggplot(descriptives, aes(age_had_dm)) + 
              geom_histogram() + 
              labs(x = "Age of diabetes diagnosis") +
              scale_x_continuous(limits = c(0, 100))
plot(agedm_histo)

#Chart of Sex
bar_sex <- ggplot(descriptives, aes(sex)) + 
                geom_bar() + 
                labs(x = "Sex of respondents") + 
                theme(axis.text.x = element_text(angle = 360))       
plot(bar_sex)

```

```{r}
summary(descriptives$age_had_dm)
```


# Model A - Logistic Regression

## Specify the model
```{r}
logistic_model <- logistic_reg() %>%
        set_engine("glm") %>%
        set_mode("classification") %>% 
        fit(ever_depression ~ ., data = train_data)
```

## Use trained model on test data
```{r}
pred_class <- predict(logistic_model,
                      new_data = test_data,
                      type = "class")

table(pred_class$.pred_class)
```

### Observed counts in outcome (test data)
```{r}
table(test_data$ever_depression)
```

### Generate predicted probabilities
```{r}
pred_prob <- predict(logistic_model,
                      new_data = test_data,
                      type = "prob")
head(pred_prob)
```

### Compare predicted probabilities to observed class
```{r}
diabetes_results <- test_data %>%
  select(ever_depression) %>%
  bind_cols(pred_class, pred_prob)

head(diabetes_results)
```

## Model Evaluation
### Confusion matrix
```{r}
conf_mat(diabetes_results, truth = ever_depression,
         estimate = .pred_class)
```

### Accuracy
```{r}
accuracy(diabetes_results, truth = ever_depression,
         estimate = .pred_class)
```

### Sensitivity
```{r}
sens(diabetes_results, truth = ever_depression,
    estimate = .pred_class)
```

### Specificity
```{r}
spec(diabetes_results, truth = ever_depression,
    estimate = .pred_class)
```

### Precision
```{r}
precision(diabetes_results, truth = ever_depression,
    estimate = .pred_class)
```

### ROC Curve
```{r}
roc_auc(diabetes_results,
        truth = ever_depression,
        .pred_0)

roc_curve <- diabetes_results %>%
  roc_curve(truth = ever_depression, .pred_0) %>%
  autoplot()

plot(roc_curve)
```

#### All metrics
```{r}
metrics <- metric_set(accuracy, sens, spec, precision, recall, f_meas)

all_metrics_lr <- metrics(diabetes_results,
               truth = ever_depression,
               estimate = .pred_class)
               
kable(all_metrics_lr)
```

### Feature importance
```{r}
coeff <- tidy(logistic_model) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(abs(estimate))) %>%
  slice_head(n = 15)

kable(coeff)

ggplot(coeff, aes(x = reorder(term, abs(estimate)), y = estimate, fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("tomato", "steelblue"),
                    labels = c("Decrease", "Increase"),
                    name   = "Odds person with DM has depression") +
  labs(
    title = "Top 10 Most Important Features — Logistic Regression",
    x     = NULL,
    y     = "Coefficient Estimate"
  ) +
  theme_minimal()
```

# Model B - K Nearest Neighbor

## Create KNN Recipe
```{r}
diabetes_recipe <- 
  recipe(ever_depression ~ ., data = train_data) %>%
  step_unknown() %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>% ### Mean center and standardize (z-score) the numeric predictors
  step_zv(all_predictors()) 
```

## Set KNN Engine
```{r}
knn_m <- nearest_neighbor(neighbors = tune()) |>
                set_mode("classification") |> 
                set_engine("kknn")

k_tuning <- grid_regular(
              neighbors(range = c(2,10)
              ))

ctrl <- control_resamples(save_pred = TRUE)

cross_validation <- vfold_cv(train_data, v = 5)
```

## Create KNN Workflow
```{r}
diabetes_wflow_knn <- 
  workflow() %>% 
  add_model(knn_m) %>% 
  add_recipe(diabetes_recipe)

diabetes_wflow_knn
```

## Tune KNN Hyperparameters
```{r}
knn_tune <- tune_grid(
              diabetes_wflow_knn, 
              resamples = cross_validation,
              grid = k_tuning, 
              control = ctrl
)

knn_tune |> collect_metrics()
```

## Use KNN on Train Data
```{r}
best_neighbors <- knn_tune |>
                    select_best(metric = "accuracy") |> 
                    pull(neighbors)


knn_best_train <- diabetes_wflow_knn |>
                    finalize_workflow(
                      select_best(x = knn_tune, metric = "accuracy")
                    )

knn_best_train
```

## Use KNN on Test Data
```{r}
knn_best_train_fit <- knn_best_train |> last_fit(data_split)

knn_best_train_fit |> collect_metrics()
```

# Session Info
```{r results='hide'}
sessionInfo()
```

