---
title: "Additional data cleaning"
author: "Richard Adejumola"
date: "2026-04-10"
output: 
  html_document:
    toc: true
    number_sections: true
    code_folding: show
---

# Import data sets
```{r}

data_myvars <- read_csv("/Users/richard/Library/CloudStorage/OneDrive-UniversityofSaskatchewan/CHEP898/finalpaper/data/brfss24_fpdata_nomiss.csv")

```


# View summary of imported data
```{r}
summary(data_myvars)
```

# Data preprocessing

At this stage, we are considering a subset of variables from the ones filtered from the original survey data. 

## Select candidate variables for the final models
```{r}
data_myvars <- data_myvars %>% 
      select(
        id, metropolis, sex, bmi_cat, age_grp, marital, education, employment,
        ethnicity, physical_health, mental_health, memory, walk, deaf, sight,
        health_rating, last_checkup, exercise, 
        heart_attack, stroke, asthma, ever_depression, ever_arthritis, age_had_dm, 
        dm_type, checked_hba1c, dm_education) 

summary(data_myvars)
```

## Clean variables 

Some levels of variables include invalid levels that should be considered missing. For instance, income, education, employment, and marital status, which all have levels "7", 9", and "11", which are missing response categories in the 2024 BRFSS survey codebook. 
```{r}

data_myvars_clean <- data_myvars %>% 
    mutate(
      education = case_when(
        education > 4 ~ NA_real_,
        .default = education
      ),
      employment = case_when(
        employment > 8 ~ NA_real_,
          .default = employment
      ),
      marital = case_when(
        marital > 6 ~ NA_real_,
          .default = marital
      ),
      physical_health = case_when(
        physical_health > 3 ~ NA_real_,
          .default = physical_health
      ),
      mental_health = case_when(
        mental_health > 3 ~ NA_real_,
          .default = mental_health
      ),
      memory = case_when(
        memory > 2 ~ NA_real_,
          .default = memory
      ),
      walk = case_when(
        walk > 2 ~ NA_real_,
          .default = walk
      ),
      deaf = case_when(
        deaf > 2 ~ NA_real_,
          .default = deaf
      ),
      sight = case_when(
        sight > 2 ~ NA_real_,
        .default = sight
      ),
      last_checkup = case_when(
        last_checkup > 4 ~ NA_real_,
          .default = last_checkup
      ),
      exercise = case_when(
        exercise > 2 ~ NA_real_,
          .default = exercise
      ),
      heart_attack = case_when(
        heart_attack > 2 ~ NA_real_,
          .default = heart_attack
      ),
      stroke = case_when(
        stroke > 2 ~ NA_real_,
          .default = stroke
      ), 
      asthma = case_when(
        asthma > 2 ~ NA_real_,
          .default = asthma
      ), 
      ever_depression = case_when(
        ever_depression > 2 ~ NA_real_,
          .default = ever_depression
      ),
      ever_arthritis = case_when(
        ever_arthritis > 2 ~ NA_real_,
          .default = ever_arthritis
      ),
      age_had_dm = case_when(
        age_had_dm > 97 ~ NA_real_,
          .default = age_had_dm
      ),
      dm_type = case_when(
        dm_type > 7 ~ NA_real_,
          .default = dm_type
      ),
      checked_hba1c = case_when(
        checked_hba1c == 88 ~ 0,
        checked_hba1c == 98 ~ 0,
        checked_hba1c > 76 ~ NA_real_,
          .default = checked_hba1c
      ),
      dm_education = case_when(
        dm_education > 8 ~ NA_real_,
          .default = dm_education
      )
      
    )

summary(data_myvars_clean)
```



## Randomly assign missing diabetes type
```{r}
data_myvars_clean %>% tabyl(dm_type)
```

Some respondents reported that they don't know their diabetes type. As Type 2 diabetes makes up approximately 90% of all types of diabetes diagnosed in the United States, we can randomly assign that proportion of unknown diabetes type to Type 2 DM and the other 10% to Type 1 DM. 

```{r}

set.seed(1123581321)

data_myvars_clean2 <- data_myvars_clean %>%
  mutate(row_id = row_number()) %>%
  
  left_join(
    data_myvars_clean %>%
      mutate(row_id = row_number()) %>%
      filter(dm_type == 7) %>%
      slice_sample(prop = 0.9) %>%
      mutate(to_recode = TRUE) %>%
      select(row_id, to_recode),
    by = "row_id"
  ) %>%
  
  mutate(
    dm_type = case_when(
      to_recode == TRUE ~ 2,      
      dm_type == 7 ~ 1,
      .default = dm_type
    )
  ) %>%
  
  select(-row_id, -to_recode)   

data_myvars_clean2 %>% tabyl(dm_type) 
```

## Review missingness in candidate variables for final model

```{r}
skim(data_myvars_clean2)
data_myvars_clean2 %>% tabyl(sex)
data_myvars_clean2 %>% tabyl(metropolis)
data_myvars_clean2 %>% tabyl(bmi_cat)
data_myvars_clean2 %>% tabyl(marital)
data_myvars_clean2 %>% tabyl(ethnicity)
data_myvars_clean2 %>% tabyl(education)
data_myvars_clean2 %>% tabyl(employment)

data_myvars_clean2 %>% tabyl(mental_health)
data_myvars_clean2 %>% tabyl(physical_health)
data_myvars_clean2 %>% tabyl(age_grp)
data_myvars_clean2 %>% tabyl(last_checkup)
data_myvars_clean2 %>% tabyl(exercise)
data_myvars_clean2 %>% tabyl(heart_attack)
data_myvars_clean2 %>% tabyl(asthma)

data_myvars_clean2 %>% tabyl(stroke)

data_myvars_clean2 %>% tabyl(ever_depression)

data_myvars_clean2 %>% tabyl(ever_arthritis)
data_myvars_clean2 %>% tabyl(dm_type)

data_myvars_clean2 %>% tabyl(dm_education)
data_myvars_clean2 %>% tabyl(memory)
data_myvars_clean2 %>% tabyl(walk)
data_myvars_clean2 %>% tabyl(deaf)
data_myvars_clean2 %>% tabyl(sight)
```

```{r}
data_myvars_clean2 <- data_myvars_clean2 %>% 
      select(
        id, sex, metropolis, bmi_cat, marital, education, employment, ethnicity, age_grp, physical_health, mental_health, last_checkup, exercise, heart_attack, stroke, sight, memory, walk, deaf, ever_arthritis, dm_education, asthma, ever_depression, age_had_dm, dm_type) 

summary(data_myvars_clean2)
```

## Drop variables with missing values
```{r}
final_data <- data_myvars_clean2 %>% 
      drop_na(
        id, sex, metropolis, bmi_cat, marital, education, employment, ethnicity, age_grp, physical_health, mental_health, last_checkup, exercise, dm_education, heart_attack, stroke, sight, memory, walk, deaf, ever_arthritis, asthma, ever_depression, age_had_dm, dm_type) 

summary(final_data)
```

## Recode reference level of variables
```{r}

final_data_recoded <- final_data %>% 
    mutate(
      sex = case_when(
        sex == 2 ~ 1,
        sex == 1 ~ 0, #male ref level
        .default = sex
      ),
      metropolis = case_when(
        metropolis == 1 ~ 0, #yes
        metropolis == 2 ~ 1, 
        .default = metropolis
      ),
      sight = case_when(
        sight == 2 ~ 0, #no ref
        .default = sight
      ),
      memory = case_when(
        memory == 2 ~ 0, #no ref
        .default = memory
      ),
      walk = case_when(
        walk == 2 ~ 0, #no ref
        .default = walk
      ),
      deaf = case_when(
        deaf == 2 ~ 0, #no ref
        .default = deaf
      ),
      bmi_cat = case_when(
        bmi_cat == 2 ~ 0, #normal weight ref level
        bmi_cat == 1 ~ 1,
        bmi_cat == 3 ~ 2,
        bmi_cat == 4 ~ 3,
        .default = bmi_cat
      ),
      marital = case_when(
        marital == 1 ~ 0, #married ref
        marital == 2 ~ 1,
        marital == 3 ~ 1,
        marital == 4 ~ 1, 
        marital == 5 ~ 2,
        marital == 6 ~ 2,
          .default = marital
      ),
      education = case_when(
        education == 4 ~ 0, #college complete ref
        .default = education
      ),
      employment = case_when(
        employment == 1 ~ 0, #employed/self-employed ref.
        employment == 2 ~ 0,
        employment == 3 ~ 1,
        employment == 4 ~ 1,
        employment == 5 ~ 1,
        employment == 6 ~ 1,
        employment == 7 ~ 1,
        employment == 8 ~ 1,
        .default = employment
      ),
       
      ethnicity = case_when(
        ethnicity == 1 ~ 0, #white ref.
        ethnicity == 2 ~ 1, #black
        ethnicity == 3 ~ 2,
        ethnicity == 4 ~ 2, 
        ethnicity == 5 ~ 2,
        ethnicity == 6 ~ 2,
          .default = ethnicity
      ),
      age_grp = case_when(
        age_grp == 6 ~ 0, #65 years and older ref
        age_grp == 1 ~ 1,
        age_grp == 2 ~ 1,
        age_grp == 3 ~ 1,
        age_grp == 4 ~ 1,
        age_grp == 5 ~ 2,
          .default = age_grp
      ),
      physical_health = case_when(
        physical_health == 1 ~ 0, #zero days ref
        physical_health == 2 ~ 1,
        physical_health == 3 ~ 2,
          .default = physical_health
      ),
      mental_health = case_when(
        mental_health == 1 ~ 0, #zero days ref
        mental_health == 2 ~ 1,
        mental_health == 3 ~ 2,
          .default = mental_health
      ),
      last_checkup = case_when(
        last_checkup == 1 ~ 0, #within last year ref
        last_checkup == 2 ~ 1, 
        last_checkup == 3 ~ 1, 
        last_checkup == 4 ~ 1,
        last_checkup == 5 ~ 1,
          .default = last_checkup
      ),
      exercise = case_when(
        exercise == 1 ~ 0, # yes ref
        exercise == 2 ~ 1, 
          .default = exercise
      ),
      stroke = case_when(
        stroke == 2 ~ 0,
          .default = stroke
      ), 
      asthma = case_when(
        asthma == 2 ~ 0,
          .default = asthma
      ), 
      
      heart_attack = case_when(
        heart_attack == 2 ~ 0,
          .default = heart_attack
      ),
      
      ever_depression = case_when(
        ever_depression == 2 ~ 0,
          .default = ever_depression
      ),
      
      ever_arthritis = case_when(
        ever_arthritis == 2 ~ 0,
          .default = ever_arthritis
      ),
     
      dm_type = case_when(
        dm_type == 2 ~ 0,
          .default = dm_type
      ),
    
  
      dm_education = case_when(
        dm_education <= 6 ~ 0,
        dm_education == 7 ~ 1,
        dm_education == 8 ~ 1,
          .default = dm_education
      )
    )

summary(final_data_recoded)
```

### Check to see that recoding worked
```{r}
skim(final_data_recoded)
final_data_recoded %>% tabyl(sex)
final_data_recoded %>% tabyl(metropolis)
final_data_recoded %>% tabyl(bmi_cat)
final_data_recoded %>% tabyl(marital)
final_data_recoded %>% tabyl(ethnicity)
final_data_recoded %>% tabyl(education)
final_data_recoded %>% tabyl(employment)

final_data_recoded %>% tabyl(mental_health)
final_data_recoded %>% tabyl(physical_health)
final_data_recoded %>% tabyl(age_grp)
final_data_recoded %>% tabyl(last_checkup)
final_data_recoded %>% tabyl(exercise)
final_data_recoded %>% tabyl(heart_attack)
final_data_recoded %>% tabyl(asthma)

final_data_recoded %>% tabyl(stroke)
final_data_recoded %>% tabyl(ever_depression)
final_data_recoded %>% tabyl(ever_arthritis)

final_data_recoded %>% tabyl(dm_type)
final_data_recoded %>% tabyl(dm_education)
final_data_recoded %>% tabyl(memory)
final_data_recoded %>% tabyl(walk)
final_data_recoded %>% tabyl(deaf)
final_data_recoded %>% tabyl(sight)
```

```{r}

rm_covsum(
  data = final_data_recoded,
  covs = c('sex', 'metropolis', 'bmi_cat', 'marital', 'education', 'employment', 'ethnicity',
        'age_grp', 'physical_health', 'mental_health', 'last_checkup', 'exercise', 'heart_attack', 'stroke', 'asthma', 'ever_depression', 'ever_arthritis', 'age_had_dm', 'dm_type', 'dm_education')
        )

```

# Export the data
```{r}
write_csv(final_data_recoded, "/Users/richard/Library/CloudStorage/OneDrive-UniversityofSaskatchewan/CHEP898/finalpaper/data/final_data_recoded.csv")
```

